Faster generic IND-CCA secure KEM using encrypt-then-MAC

### 2024-10-12 etm-transform/manuscript/main.tex
Revised manuscirpt in accordance to suggestions

@@

%%%% 6. ABSTRACT %%%%
\begin{abstract}
    Indistinguishability under chosen-ciphertext attack (IND-CCA) is the accepted standard of security for key encapsulation mechanisms (KEM). However, provably chosen-ciphertext security is difficult to achieve from scratch. In this work, we generalize the ``Hashed ElGamal'' encryption scheme proposed by M. Abdalla et al into an IND-CCA secure KEM by combining an OW-PCA secure public-key encryption (PKE) scheme with an existentially unforgeable message authentication code (MAC) in a pattern called encrypt-then-MAC. Compared to the Fujisaki-Okamoto transformation, the encrypt-then-MAC transformation replaces de-randomization and re-encryption with computing a MAC tag. When instantiated with the PKE sub-routines of ML-KEM, the encrypt-then-MAC transformation achieves massive computational efficiency improvements over ML-KEM in a variety of practical scenarios.
\end{abstract}

@@

\subsection{Our contribution}
In this paper, we propose a generic IND-CCA secure KEM using techniques adapted from DHIES. We call our proposed scheme the encrypt-then-MAC transformation due to the conceptual similarity to the namesake symmetric encryption technique for achieving authenticated encryption. Our scheme mainly differs from DHIES in its versatility and input requirement. Whereas the IND-CCA security of DHIES reduces specifically to the Gap Diffie-Hellman assumption, the chosen-ciphertext security of the encrypt-then-MAC KEM reduces more generally to the OW-PCA security \cite{DBLP:conf/ctrsa/OkamotoP01} of the input scheme (we will show in Section \ref{sec:main-results} that the Gap Diffie-Hellman assumption implies OW-PCA security). In addition, we propose that because each call to encapsulation samples a fresh PKE plaintext, the encrypt-then-MAC KEM can be instantiated with one-time secure MAC such as Poly1305 (M. Abdalla et al. originally proposed to use HMAC and CBC-MAC, which are many-time secure MAC but less efficient than one-time MAC, see Section \ref{sec:choosing-a-message-authenticator}), which greatly improves the computational efficiency of the scheme.

\begin{figure}
    \centering
    \begin{tikzpicture}
        \tikzstyle{rect} = [draw, rectangle, minimum width=3cm, minimum height=1cm]
        \tikzstyle{filledrect} = [draw, rectangle, minimum width=3cm, minimum height=1cm, fill=lightgray]
        \matrix [column sep=7mm, row sep=5mm] {
            \node (pk) [rect] {$\pk$}; &
            \node (m) [rect] {$m \leftsample \mathcal{M}$}; &
            \node (hashg) [rect] {Hash function $G$}; \\
            \node (pkeenc) [rect] {$\texttt{PKE.Enc}$}; &
            \node (mackey) [rect] {$k_\mac$}; \\
            \node (ct) [filledrect] {$c^\prime$}; &
            \node (mac) [rect] {$\mac$}; &
            \node (tag) [filledrect] {$t$}; \\
        };
        \draw[->, thick] (pk) -- (pkeenc);
        \draw[->, thick] (m) -- (pkeenc);
        \draw[->, thick] (m) -- (hashg);
        \draw[->, thick] (hashg) -- (mackey);
        \draw[->, thick] (pkeenc) -- (ct);
        \draw[->, thick] (mackey) -- (mac);
        \draw[->, thick] (ct) -- (mac);
        \draw[->, thick] (mac) -- (tag);
    \end{tikzpicture}

    \caption{Combining PKE with MAC using encrypt-then-MAC to encapsulate a shared secret. 
    The returned values are colored grey
    }\label{fig:etm-diagram}
\end{figure}

\textbf{Construction.} From a high level, the encrypt-then-MAC transformation takes as input a PKE and a MAC, then outputs a KEM. The output KEM encapsulates by encrypting a randomly sampled PKE plaintext and hashing the PKE plaintext-ciphertext pair into the shared secret. A symmetric key is derived from hashing the PKE plaintext, then used to produce a MAC tag on the PKE ciphertext. At decapsulation, the PKE secret key is used to recover the PKE plaintext, which is then hashed into a symmetric key that is used to verify the MAC tag against the input ciphertext. A summary of the data flow in the encrypt-then-MAC transformation can be found in Figure \ref{fig:etm-diagram}. A detailed description of individual routines can be found in Figure \ref{fig:etm-routines}.

\textbf{Security.} In Section \ref{sec:main-results}, we prove that if the input PKE is one-way secure against plaintext-checking attack (OW-PCA) and the MAC is one-time existentially unforgeable, then the output KEM is IND-CCA secure. The IND-CCA security of the output KEM reduces tightly to the OW-PCA security of the input PKE, and non-tightly to the unforgeability of the MAC, though the non-tightness will likely not be a weakness because the construction makes it easy to plug in MAC of any desired security level.

\textbf{Implementation and performance comparison.} In Section \ref{sec:practical-instantiation-with-mlkem} and \ref{sec:performance-comparison}, we applied the encrypt-then-MAC transformation to the K-PKE subroutines defined in ML-KEM \cite{FIPS203}, which we call $\mlkemplus$. Compared with ML-KEM, which is based on the $U^{\not\bot}$ variant (implicit rejection with rigidity) of the modular Fujisaki-Okamoto KEM transformation, $\mlkemplus$ demonstrates massive performance improvements in decapsulation while only incurring minimal performance penalty in encapsulation runtime and ciphertext size. On an AMD EPYC 9R14 CPU, $\mlkemplus$ with Poly1305 achieves between 72\%-80\% reduction of CPU cycles needed for decapsulation while only incurring 2\%-7\% increase of CPU cycles needed for encapsulation. See Table \ref{tbl:cpu-cycles-summary} for a summary of performance comparison of individual routines against ML-KEM.

@@

We also measured the round trip time of key exchange protocols with various modes of authentication. When compared to ML-KEM, $\mlkemplus$ achieves 24\%-28\% reduction of round trip time in unauthenticated key exchange (KE), 29\%-35\% in unilaterally authenticated key exchange (UAKE), and 35\%-48\% reduction in mutually authenticated key exchange (AKE). See Table \ref{tbl:rtt-summary} for a summary of round trip times.

@@

\begin{figure}
    \centering

@@

\section{Preliminaries}

@@

\section{The encrypt-then-MAC transformation}\label{sec:main-results}
In this section we introduce the encrypt-then-MAC transformation that transforms an OW-PCA secure PKE and an one-time existentially unforgeable MAC into an IND-CCA secure KEM. In Section \ref{sec:proof-of-theorem} we reduce the IND-CCA security of the KEM tightly to the OW-PCA security of the underlying PKE, and non-tightly to the unforgeability of the MAC. In Section \ref{sec:elgamal-is-ow-pca}, we show that the ``Hashed ElGamal'' cryptosystem in \cite{DBLP:conf/ctrsa/AbdallaBR01} is a special case of the encrypt-then-MAC transformation by reducing the OW-PCA security of the ElGamal cryptosystem to the Gap Diffie-Hellman assumption.

@@

The key generation routine of $\kem_\etm$ is largely identical to that of the $\pke$, only a secret value $z$ is sampled as the implicit rejection symbol. In the encapsulation routine, a MAC key is derived from the randomly sampled plaintext $k \leftarrow G(m)$, then used to sign the unauthenticated ciphertext $c^\prime$. Because the encryption routine might be randomized, the session key is derived from both the message and the ciphertext. Finally, the unauthenticated ciphertext $c^\prime$ and the tag $t$ combine into the authenticated ciphertext $c$ that would be transmitted to the peer. In the decapsulation routine, the decryption $\hat{m}$ of the unauthenticated ciphertext is used to re-derive the MAC key $\hat{k}$, which is then used to re-compute the tag $\hat{t}$. The ciphertext is considered valid if and only if the recomputed tag is identical to the input tag.

For an adversary $A$ to produce a valid tag $t$ for some unauthenticated ciphertext $c^\prime$ under the symmetric key $k \leftarrow G(\decrypt(\sk^\prime, c^\prime))$ implies that $A$ must either know the symmetric key $k$ or produce a forgery. Under the random oracle model, $A$ also cannot know $k$ without knowing its preimage $\decrypt(\sk^\prime, c^\prime)$, so $A$ must either have produced $c^\prime$ honestly, or have broken the one-way security of $\pke$. This means that the decapsulation oracle will not give out information on decryptions that the adversary does not already know.

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{0.6\textwidth}
    \begin{algorithm}[H]
        \caption*{$\pco(m, c)$}
        \begin{algorithmic}[1]
            \State $k \leftarrow G(m)$
            \State $t \leftarrow \mac(k, c)$
            \State \Return $\llbrack \mathcal{O}^\decap((c, t)) = H(m, c)\rrbrack$
        \end{algorithmic}
    \end{algorithm}
    \end{minipage}
    \caption{With the encrypt-then-MAC transformation, every chosen-ciphertext attack against the KEM can be converted into a plaintext-checking attack against the underlying PKE}\label{fig:pco-using-decap-oracle}
\end{figure}

@@

\section{Practical instantiation with ML-KEM}\label{sec:practical-instantiation-with-mlkem}
ML-KEM is an IND-CCA secure key encapsulation mechanism standardized by NIST in FIPS 203 \cite{FIPS203}. The chosen-ciphertext security of ML-KEM is achieved in two steps. First, ML-KEM constructs a PKE whose IND-CPA security reduces to the conjectured intractability of the decisional Module Learning with Error (MLWE) problem. Then, the $U^{\not\bot}_m$ variant of the Fujisaki-Okamoto transformation is applied to convert the IND-CPA secure PKE into an IND-CCA secure KEM. A summary of the IND-CPA PKE routines can be found in Figure \ref{fig:k-pke-routines}. We refer readers to \cite{DBLP:conf/tcc/HofheinzHK17} and \cite{FIPS203} for more details of the $U^{\not\bot}_m$ variant of the Fujisaki-Okamoto transformation.

@@

We applied the encrypt-then-MAC transformation to the K-PKE routines of ML-KEM. The resulting KEM is called $\mlkemplus$. A complete description of the KEM routines can be found in Figure \ref{fig:ml-kem-plus-routines}

@@

    \begin{minipage}[t]{0.5\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\keygen\texttt{()}$}
            \begin{algorithmic}[1]
                \State $z \leftsample \{0,1\}^{256}$
                \State $(\pk, \sk^\prime) \leftsample \texttt{K-PKE}.\keygen()$
                \State $h \leftarrow H(\pk)$
                \State $\sk \leftarrow (sk^\prime \Vert \pk \Vert h \Vert z)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}\vspace{-0.5cm}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\encap\texttt{(\pk)}$}
            \begin{algorithmic}[1]
                \Require Public key $\pk$
                \State $m \leftsample \{0,1\}^{256}$
                \State $(\overline{K}, k) \leftarrow G(m \Vert H(\pk))$
                \State $c^\prime \leftsample \texttt{K-PKE}.\encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c^\prime)$
                \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert c^\prime)$
                \State $c \leftarrow (c^\prime, t)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\decap\texttt{(\sk, c)}$}
            \begin{algorithmic}[1]
                \Require Secret key $\sk = (\sk^\prime \Vert \pk \Vert h \Vert z)$
                \Require Ciphertext $c = (c^\prime \Vert t)$
                \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
                \State $(c^\prime, t) \leftarrow c$
                \State $\hat{m} \leftarrow \texttt{K-PKE}.\decrypt(\sk^\prime, c^\prime)$
                \State $(\overline{K}, \hat{k}) \leftarrow G(\hat{m} \Vert h)$
                \State $\hat{t} \leftarrow \mac(\hat{k}, c^\prime)$
                \If{$\hat{t} = t$}
                    \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \Else
                    \State $K \leftarrow \texttt{KDF}(z \Vert t)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}

    \caption{$\texttt{ML-KEM}^+$ applies our encrypt-then-MAC transformation to K-PKE}\label{fig:ml-kem-plus-routines}
\end{figure}

@@

\subsection{Choosing a message authenticator}\label{sec:choosing-a-message-authenticator}
For implementation, we instantiated the MAC with a selection that covered a wide range of MAC designs, including Poly1305 \cite{bernstein2005poly1305}, GMAC \cite{mcgrew2004galois}, CMAC \cite{iwata2003omac}\cite{black2000cbc}, and KMAC \cite{SP80053r4}.

@@

\begin{table}[H]
    \centering
    \footnotesize
    \caption{CPU cycles needed to compute tag on various input sizes}\label{tbl:mac-performance}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 768 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 909 & 2823 \\
        \hline
        GMAC & 3899 & 4859 \\
        \hline
        CMAC & 6291 & 6373 \\
        \hline
        KMAC & 6373 & 7791 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1088 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 961 & 2704 \\
        \hline
        GMAC & 3899 & 4827 \\
        \hline
        CMAC & 7305 & 7588 \\
        \hline
        KMAC & 9697 & 9928 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1568 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 1065 & 1809 \\
        \hline
        GMAC & 4055 & 5026 \\
        \hline
        CMAC & 8735 & 8772 \\
        \hline
        KMAC & 11647 & 12186 \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{table}[H]
    \centering
    \footnotesize
    \caption{CPU cycles of each KEM routine}\label{tbl:kem-performance}

@@

\subsection{Key exchange protocols}\label{sec:key-exchange-protocols}
A common application of key encapsulation mechanism is key exchange protocols, where two parties establish a shared secret using a public channel. We used $\mlkemplus$ to implement the key exchange protocols described in \cite{DBLP:conf/eurosp/BosDKLLSSSS18} and compared the round trip time (RTT) of $\mlkemplus$ key exchange with ML-KEM key exchange. We denote the party who sends the first message to be the client and the other party to be the server. Round trip time (RTT) is defined to be the time interval between the moment before the client starts generating ephemeral keypairs and the moment after the client derives the final session key. All experiements are run on a pair of AWS c7a.medium instances both located in the \texttt{us-west-2} region. For each experiment, a total of 10,000 rounds of key exchange are performed, with the median and average round trip time (measured in microsecond) recorded.

>>>>

% TODO: describe ML-KEM in appendix
%   Added in Figure 14

% TODO: why is key derived from H(preKey || c') vs H(preKey || t)
% DONE: that was a typo and was fixed. the implementation hashed the tag, not the ciphertext

% TODO: Decapsulation line 9
% DONE: ML-KEM still uses implicit rejection, see line 7 of Algorithm 18 in FIPS 203 (https://nvlpubs.nist.gov/nistpubs/FIPS/NIST.FIPS.203.pdf)

% TODO: Explain the removal of the pseudorandom coin
% DONE: added paragraph in section 3: "Since the encrypt-then-MAC transformation removes re-encryption..."

% TODO: reference for key exchange protocols
% DONE: Added to the beginning of section 5.3

% TODO: revise abstract and introduction
% DONE: revised abstract, reorganized introduction, removed RSA/ElGamal in the last section

% TODO: Why only 128-bit tag?
% DONE: Added a few sentences in section 4.1

@@

%%%% 6. ABSTRACT %%%%
\begin{abstract}
    ML-KEM is a lattice-based IND-CCA secure key encapsulation mechanism (KEM) standardized by NIST in FIPS 203. ML-KEM achieves chosen-ciphertext security by applying the Fujisaki-Okamoto transformation to an IND-CPA secure public-key encryption (PKE) scheme. The Fujisaki-Okamoto transformation uses de-randomization and re-encryption to ensure ciphertext non-malleability, but because ML-KEM's underlying PKE encryption is significantly slower than the underlying PKE decryption, ML-KEM's decapsulation performance is dominated by re-encryption. In this paper, we propose an alternative generic IND-CCA secure KEM transformation that applies the encrypt-then-MAC pattern to an OW-PCA secure PKE and an existentially unforgeable MAC. Compared to the Fujisaki-Okamoto transformation, our encrypt-then-MAC transformation replaces de-randomization and re-encryption with computing a MAC tag. We then implemented $\mlkemplus$ by applying the encrypt-then-MAC transformation to the PKE sub-routines of ML-KEM and a wide selection of MACs including Poly1305, GMAC, CMAC, and KMAC. At all three security levels (128-bit, 196-bit, and 256-bit security), our implementation demonstrates great performance improvements in a variety of practical scenarios. For example, $\mlkemplus$ with Poly1305 reduces decapsulation's CPU cycles by 72\%-79\% and decreases network round trip time in key exchange protocols by 24\%-48\%.
\end{abstract}

@@

\subsection{Our contribution}
\paragraph{Encrypt-then-MAC: a generic IND-CCA secure KEM transformation} In this paper, we propose a generic IND-CCA secure KEM using techniques adapted from DHIES. We call our proposed scheme the encrypt-then-MAC transformation due to the conceptual similarity to the namesake symmetric encryption technique for achieving authenticated encryption. We prove that the IND-CCA security of the encrypt-then-MAC transformation reduces tightly to the OW-PCA security of the underlying PKE and the existential unforgeability of the underlying MAC. In addition, we argue that the encrypt-then-MAC transformation can be instantiated with one-time MAC for further performance improvements.

\paragraph{$\mlkemplus$: an instantiation of encrypt-then-MAC transformation with ML-KEM's PKE sub-routines.} We implemented $\mlkemplus$ by applying our encrypt-then-MAC transformation to ML-KEM's PKE sub-routines. Compared with ML-KEM, which is based on the $U^{\not\bot}$ variant (implicit rejection with rigidity) of the modular Fujisaki-Okamoto KEM transformation, $\mlkemplus$ demonstrates massive performance improvements in decapsulation while only incurring minimal performance penalty in encapsulation runtime and ciphertext size. On an AMD EPYC 9R14 CPU, $\mlkemplus$ with Poly1305 achieves between 72\%-80\% reduction of CPU cycles needed for decapsulation while only incurring 2\%-7\% increase of CPU cycles needed for encapsulation (see Table \ref{tbl:cpu-cycles-summary}).

@@

\paragraph{Performance improvements in practical key exchange.} We also implemented the three Kyber key exchange protocols described in \cite{DBLP:conf/eurosp/BosDKLLSSSS18} using $\mlkemplus$ and ML-KEM. Compared to ML-KEM, $\mlkemplus$ achieves 24\%-28\% reduction of round trip time in unauthenticated key exchange (KE), 29\%-35\% in unilaterally authenticated key exchange (UAKE), and 35\%-48\% reduction in mutually authenticated key exchange (AKE). See Table \ref{tbl:rtt-summary} for a summary of round trip times.

@@

@@

The rest of the paper is organized as follows. In Section \ref{sec:preliminaries}, we define and discuss the preliminary concepts and theorems. In Section \ref{sec:main-results}, we present our encrypt-then-MAC transformation in details, prove its security results, and discusses its relationship to the ElGamal cryptosystem. In Section \ref{sec:practical-instantiation-with-mlkem}, we present $\mlkemplus$, an instantiation of the encrypt-then-MAC transformation using ML-KEM's PKE sub-routines, then discuss some implementation rationales. In Section \ref{sec:performance-comparison}, we benchmark and compare the performance of individual KEM routines and key exchange round trip times between $\mlkemplus$ and ML-KEM.

\section{Preliminaries}\label{sec:preliminaries}

@@

\section{The encrypt-then-MAC transformation}\label{sec:main-results}
In this section we introduce the encrypt-then-MAC transformation that transforms an OW-PCA secure PKE and an one-time existentially unforgeable MAC into an IND-CCA secure KEM. Our scheme mainly differs from DHIES in its versatility and input requirement. Whereas the IND-CCA security of DHIES reduces specifically to the Gap Diffie-Hellman assumption, the chosen-ciphertext security of the encrypt-then-MAC KEM reduces more generally to the OW-PCA security \cite{DBLP:conf/ctrsa/OkamotoP01} of the input scheme (we will show in Section \ref{sec:elgamal-is-ow-pca} that the Gap Diffie-Hellman assumption implies OW-PCA security). In addition, we propose that because each call to encapsulation samples a fresh PKE plaintext, the encrypt-then-MAC KEM can be instantiated with one-time secure MAC such as Poly1305 (M. Abdalla et al. originally proposed to use HMAC and CBC-MAC, which are many-time secure MAC but less efficient than one-time MAC, see Section \ref{sec:choosing-a-message-authenticator}), which greatly improves the computational efficiency of the scheme. A summary of encapsulation data flow is illustrated in Figure \ref{fig:etm-diagram}.

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
        \tikzstyle{rect} = [draw, rectangle, minimum width=3cm, minimum height=1cm]
        \tikzstyle{filledrect} = [draw, rectangle, minimum width=3cm, minimum height=1cm, fill=lightgray]
        \matrix [column sep=7mm, row sep=5mm] {
            \node (pk) [rect] {$\pk$}; &
            \node (m) [rect] {$m \leftsample \mathcal{M}$}; &
            \node (hashg) [rect] {Hash function $G$}; \\
            \node (pkeenc) [rect] {$\texttt{PKE.Enc}$}; &
            \node (mackey) [rect] {$k_\mac$}; \\
            \node (ct) [filledrect] {$c^\prime$}; &
            \node (mac) [rect] {$\mac$}; &
            \node (tag) [filledrect] {$t$}; \\
        };
        \draw[->, thick] (pk) -- (pkeenc);
        \draw[->, thick] (m) -- (pkeenc);
        \draw[->, thick] (m) -- (hashg);
        \draw[->, thick] (hashg) -- (mackey);
        \draw[->, thick] (pkeenc) -- (ct);
        \draw[->, thick] (mackey) -- (mac);
        \draw[->, thick] (ct) -- (mac);
        \draw[->, thick] (mac) -- (tag);
    \end{tikzpicture}

    \caption{Combining PKE with MAC using encrypt-then-MAC to ensure ciphertext integrity}\label{fig:etm-diagram}
\end{figure}

In Section \ref{sec:proof-of-theorem} we reduce the IND-CCA security of the KEM tightly to the OW-PCA security of the underlying PKE, and non-tightly to the unforgeability of the MAC. In Section \ref{sec:elgamal-is-ow-pca}, we show that the ``Hashed ElGamal'' cryptosystem in \cite{DBLP:conf/ctrsa/AbdallaBR01} is a special case of the encrypt-then-MAC transformation by reducing the OW-PCA security of the ElGamal cryptosystem to the Gap Diffie-Hellman assumption.

@@

The key generation routine of $\kem_\etm$ is largely identical to that of the $\pke$, only a secret value $z$ is sampled as the implicit rejection symbol. In the encapsulation routine, a MAC key is derived from the randomly sampled plaintext $k \leftarrow G(m)$, then used to sign the unauthenticated ciphertext $c^\prime$. Finally, the unauthenticated ciphertext $c^\prime$ and the tag $t$ combine into the authenticated ciphertext $c$ that would be transmitted to the peer. In the decapsulation routine, the decryption $\hat{m}$ of the unauthenticated ciphertext is used to re-derive the MAC key $\hat{k}$, which is then used to re-compute the tag $\hat{t}$. The ciphertext is considered valid if and only if the recomputed tag is identical to the input tag.

Since the encrypt-then-MAC transformation removes re-encryption in decapsulation, there is no longer the need for fixing the pseudorandom coin $r$ in the PKE's encryption routine. If the input PKE is already rigid, then the shared secret may be derived from hashing the PKE plaintext alone. However, if the input PKE is not rigid, then the shared secret must be derived from hashing both the PKE plaintext and the PKE ciphertext.

The chosen-ciphertext security of the encrypt-then-MAC scheme can be intuitively argued through an adversary's inability to learn additional information from the decapsulation oracle. For an adversary $A$ to produce a valid tag $t$ for some unauthenticated ciphertext $c^\prime$ under the symmetric key $k \leftarrow G(\decrypt(\sk^\prime, c^\prime))$ implies that $A$ must either know the symmetric key $k$ or produce a forgery. Under the random oracle model, $A$ also cannot know $k$ without knowing its preimage $\decrypt(\sk^\prime, c^\prime)$, so $A$ must either have produced $c^\prime$ honestly, or have broken the one-way security of $\pke$. This means that the decapsulation oracle will not give out information on decryptions that the adversary does not already know.

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{0.45\textwidth}
    \begin{algorithm}[H]
        \caption*{$\pco(m, c)$}
        \begin{algorithmic}[1]
            \State $k \leftarrow G(m)$
            \State $t \leftarrow \mac(k, c)$
            \State \Return $\llbrack \mathcal{O}^\decap((c, t)) = H(m, c)\rrbrack$
        \end{algorithmic}
    \end{algorithm}
    \end{minipage}
    \caption{With the encrypt-then-MAC transformation, every chosen-ciphertext attack against the KEM implies a plaintext-checking attack against the underlying PKE}\label{fig:pco-using-decap-oracle}
\end{figure}

@@

\section{Practical instantiation with ML-KEM}\label{sec:practical-instantiation-with-mlkem}
ML-KEM is an IND-CCA secure key encapsulation mechanism standardized by NIST in FIPS 203 \cite{FIPS203}. The chosen-ciphertext security of ML-KEM is achieved in two steps. First, ML-KEM constructs a PKE (called K-PKE, see Figure \ref{fig:k-pke-routines}) whose IND-CPA security reduces to the conjectured intractability of the decisional Module Learning with Error (MLWE) problem. Then, the $U^{\not\bot}_m$ variant of the Fujisaki-Okamoto transformation is applied to convert the IND-CPA secure PKE into an IND-CCA secure KEM (see Figure \ref{fig:ml-kem-routines}).

@@

    \begin{minipage}[t]{0.4\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM.KeyGen()}$}
            \begin{algorithmic}[1]
                \State $(\pk, \sk^\prime) \leftsample \texttt{K-PKE.KeyGen()}$
                \State $h \leftarrow \texttt{SHA3-256}(\pk)$
                \State $z \leftsample \mathcal{M}$
                \State $\sk \leftarrow (\sk^\prime, \pk, h, z)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM.Encap}(\pk)$}
            \begin{algorithmic}[1]
                \State $m \leftsample \mathcal{M}$
                \State $h \leftarrow \texttt{SHA3-256}(\pk)$
                \State $(K, r) \leftarrow \texttt{SHA3-512}(m, h)$
                \State $c \leftarrow \texttt{K-PKE.Enc}(\pk, m ,r)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \begin{minipage}[t]{0.4\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM.Decap}(\sk, c)$}
            \begin{algorithmic}[1]
                \State $\hat{m} \leftarrow \texttt{K-PKE.Dec}(\sk^\prime, c)$
                \State $(\overline{K}, \hat{r}) \leftarrow \texttt{SHA3-512}(\hat{m}, h)$
                \State $\hat{c} \leftarrow \texttt{K-PKE.Enc}(\pk, m, \hat{r})$
                \If{$\hat{c} = c$}
                    \State $K \leftarrow \overline{K}$
                \Else 
                    \State $K \leftarrow \texttt{SHA3-256}(c, z)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}

    \caption{ML-KEM uses the $U^{\not\bot}_m$ variant of the modular Fujisaki-Okamoto KEM transformation}\label{fig:ml-kem-routines}
\end{figure}

We applied the encrypt-then-MAC transformation to the K-PKE routines of ML-KEM. The resulting KEM is called $\mlkemplus$ (See Figure \ref{fig:ml-kem-plus-routines}).

@@

    \begin{minipage}[t]{0.5\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\keygen\texttt{()}$}
            \begin{algorithmic}[1]
                \State $z \leftsample \{0,1\}^{256}$
                \State $(\pk, \sk^\prime) \leftsample \texttt{K-PKE}.\keygen()$
                \State $h \leftarrow H(\pk)$
                \State $\sk \leftarrow (sk^\prime \Vert \pk \Vert h \Vert z)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}\vspace{-0.5cm}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\encap\texttt{(\pk)}$}
            \begin{algorithmic}[1]
                \State $m \leftsample \{0,1\}^{256}$
                \State $(\overline{K}, k) \leftarrow G(m \Vert H(\pk))$
                \State $c^\prime \leftsample \texttt{K-PKE}.\encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c^\prime)$
                \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \State $c \leftarrow (c^\prime, t)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{$\texttt{ML-KEM}^\texttt{+}\texttt{.}\decap\texttt{(\sk, c)}$}
            \begin{algorithmic}[1]
                \Require Secret key $\sk = (\sk^\prime \Vert \pk \Vert h \Vert z)$
                \Require Ciphertext $c = (c^\prime \Vert t)$
                \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
                \State $(c^\prime, t) \leftarrow c$
                \State $\hat{m} \leftarrow \texttt{K-PKE}.\decrypt(\sk^\prime, c^\prime)$
                \State $(\overline{K}, \hat{k}) \leftarrow G(\hat{m} \Vert h)$
                \State $\hat{t} \leftarrow \mac(\hat{k}, c^\prime)$
                \If{$\hat{t} = t$}
                    \State $K \leftarrow \texttt{KDF}(\overline{K} \Vert t)$
                \Else
                    \State $K \leftarrow \texttt{KDF}(z \Vert t)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}

    \caption{$\texttt{ML-KEM}^+$ applies our encrypt-then-MAC transformation to K-PKE. $G$ is SHA3-512, $H$ is SHA3-256, $\texttt{KDF}$ is Shake256}\label{fig:ml-kem-plus-routines}
\end{figure}

@@

We also modified how the shared secret is derived: instead of hashing the entire ciphertext, we hashed the MAC tag, which essentially functions as a hash of the ciphertext and thus serves the same purpose. On the other hand, because ML-KEM ciphertext sizes range from 768 bytes to 1568 bytes while the tag size is fixed at 16 bytes, applying the KDF to the tag instead of the entire ciphertext further yields meaningful performance improvements.

\subsection{Choosing a message authenticator}\label{sec:choosing-a-message-authenticator}
For implementation, we instantiated the MAC with a selection that covered a wide range of MAC designs, including Poly1305 \cite{bernstein2005poly1305}, GMAC \cite{mcgrew2004galois}, CMAC \cite{iwata2003omac}\cite{black2000cbc}, and KMAC \cite{SP80053r4}. All chosen MACs are parameterized with 256-bit keys and 128-bit tags. We chose 256-bit keys for all three security levels (128-bit, 196-bit, and 256-bit security) because choosing a smaller key size does not yield meaningful performance gains (in addition, Poly1305 only supports 256-bit keys). On the other hand, a 128-bit tag size is sufficient for all three security levels because unlike brute-force attacks on the MAC key, which can be carried out offline with large computing capacity, attacks on the tag require interaction with an online peer that plays the role of a MAC verification oracle. This means that the attack bandwidth is limited by the throughput of the online oracle instead of the attacker's computing power, and in practice it is easy to implement rate limiting in the cryptographic protocol such that the adversary cannot submit large quantities of MAC verification queries.

@@

\begin{table}[h]
    \centering
    \footnotesize
    \caption{CPU cycles needed to compute tag on various input sizes}\label{tbl:mac-performance}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 768 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 909 & 2823 \\
        \hline
        GMAC & 3899 & 4859 \\
        \hline
        CMAC & 6291 & 6373 \\
        \hline
        KMAC & 6373 & 7791 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1088 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 961 & 2704 \\
        \hline
        GMAC & 3899 & 4827 \\
        \hline
        CMAC & 7305 & 7588 \\
        \hline
        KMAC & 9697 & 9928 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1568 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 1065 & 1809 \\
        \hline
        GMAC & 4055 & 5026 \\
        \hline
        CMAC & 8735 & 8772 \\
        \hline
        KMAC & 11647 & 12186 \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{table}[h]
    \centering
    \footnotesize
    \caption{CPU cycles of each KEM routine}\label{tbl:kem-performance}

@@

\subsection{Key exchange protocols}\label{sec:key-exchange-protocols}
The CRYSTALS-Kyber team \cite{DBLP:conf/eurosp/BosDKLLSSSS18} proposed three key exchange protocols: $\texttt{Kyber.KE}$ for a passively secure ephemeral key exchange, $\texttt{Kyber.UAKE}$ for unilaterally authenticated key exchange, and \texttt{Kyber.AKE} for mutually authenticated key exchange. First formally proposed in \cite{DBLP:conf/stoc/BellareCK98} and later analyzed in \cite{DBLP:conf/eurocrypt/CanettiK01}, these key exchange protocols all achieve weak forward secrecy, and the authenticated key exchange protocols are resistant to certain categories of active adversaries (e.g. Man-in-the-Middle attacks).
